---
title: 'MDGS 2021: Data Import & Analysis Documentation'
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r urban institute data, include = FALSE}
# if necessary
# install.packages('devtools') 
# if necessary
# devtools::install_github('UrbanInstitute/education-data-package-r')

```

```{r packages, include = FALSE}
library(readr)
library(tidyverse)
library(skimr)
library(knitr)
library(janitor)
library(educationdata)
library(flextable)

```

# Introduction

## Context

The Maryland Green Schools Program (MDGS) is a sustainable schools certification program, and the signature program of the Maryland Association for Environmental and Outdoor Education (MAEOE). The MDGS program provides infrastructure, support, and a rigorous review process to certify any school in Maryland with sustainable school certification as a Maryland Green School. The MDGS program has been essential to Maryland's ability to connect with goals of the Chesapeake Bay Watershed Agreement, as well as helping schools achieve Maryland's educational standards and graduation requirements for environmental literacy. Currently, nearly 30% of all Maryland schools (public and private) are certified Green Schools. In 2020, MAEOE received state funding to expand their efforts, with a goal of having 50% of all schools in the state certified by 2026.

This document is an internal guide and explanation of the data cleaning and analysis steps undertaken and decisions made as an **internal record and guide for the JSC** team in replicating this work.

# Baseline Analysis

The first stage of analysis was to gather, wrangle, and descriptively analyze the baseline data about all schools in the state of Maryland -- public and private. This is the full population of *potential* Green Schools, to which we will comparatively analyze the current certified Green Schools.

## Data Loading

The first data we dealt with was the data from NCES about all schools in the state of Maryland - public and private. Data about public and private schools are collected separately, so there were two files that needed to be loaded, cleaned, and merged.

The public data set can be loaded already mostly cleaned from the Urban Institute.

```{r NCES load public data, include = FALSE}

# Pulling the latest data directly from the Urban Institute's API
# They have compiled/cleaned the NCES data from the CCD survey

# First code pulls down the basic school info (what I was using previously)
# fips = 24 is flitering it just for Maryland
# update the year with new data when they have it

### Future years: may need to come up wiht a system to not overwrite/lose closed schools...

all_public_dir = get_education_data(level = "schools",
                                source = "ccd",
                                topic = "directory",
                                filters = list(year = 2018,
                                               fips = 24),
                                add_labels = TRUE)

# Save this file to the "raw data" file as a CSV
# This is a backup in case it ever became difficult to access UI's data

# write.csv(all_public_dir, "raw-data/all_public_dir_040221.csv")

```

Urban Institute has data available on enrollment by race. It's "tall data," so I still need to figure out how to merge it properly.

```{r NCES load public race data, include = FALSE}
# STILL TO FIGURE OUT

# This is supposed to pull down enrollment by race/ethnicity
# It produces a "tall" data set
# Need to figure out how to do the totals/merging with my original if want to get race data into the main dataset
## Commenting out until I'm able to do that

#all_public_enroll = get_education_data(level = "schools",
#                                       source = "ccd",
#                                       topic = "enrollment",
#                                       subtopic = list("race"),
#                                       filters = list(year = 2018,
#                                                      fips = 24),
#                                       add_labels = TRUE)
  
```

Private school data seems to still need to be manually loaded from NCES download.

```{r NCES load private data, include = FALSE}

### PRIVATE SCHOOL DATA

# Import data
### These still need to be manually downloaded from NCES site
### Urban Institute doesn't seem to maintain them

# Fixed some of the variables that came through weirdly.  
# Cleaned names

all_pvt = read_csv("raw-data/nces_private_md.csv",
                   col_types = cols(
                     PSS_PHONE = col_character(),
                     PSS_LIBRARY = col_factor()))

all_pvt = clean_names(all_pvt)
```

## Data Cleaning

The Urban Institute data is cleaner than the raw source, but some variables still need to be cleaned for our analysis purposes.

### Coding Grade Bands

Although the Urban Institute set has done a pre-code of school_level, it doesn't fit exactly with how I coded it for public/private. For now I'm keeping my distinctions.

The first cleaning task was to re-code each school's grade-level data into categorical grade bands:

-   ES (includes \~7 pre-k schools)
-   ES/MS
-   MS
-   MS/HS
-   HS
-   All Grades
-   Ungraded / Not-Reporting / Other

*The definitions for the grade bands is in the supporting Word Doc data log. This document provides consistent decisions for schools that weirdly span the grade bands. The code below is structured to apply those rules.*

```{r NCES public grade levels, include = FALSE}

# Clean up the Grade Band data in each file separately

# Public Schools dataset has a field for Lowest Grade and Highest Grade
# We use those to code into grade bands


all_public = all_public_dir %>% 
  mutate(grade_band = 
           case_when(
            highest_grade_offered %in% c("Pre-K", "Kindergarten", 1, 2, 3, 4, 5, 6) ~ "ES",
            lowest_grade_offered %in% c(9,10,11,12) ~ "HS",
      lowest_grade_offered == "Kindergarten" & highest_grade_offered == 7 ~ "ES/MS",
      lowest_grade_offered == "Pre-K" & highest_grade_offered == 7 ~ "ES/MS",
      lowest_grade_offered == "Pre-K" & highest_grade_offered == 8 ~ "ES/MS",
      lowest_grade_offered == "Kindergarten" & highest_grade_offered == 8 ~ "ES/MS",
      lowest_grade_offered == 5 & highest_grade_offered == 8 ~ "MS",
      lowest_grade_offered == 6 & highest_grade_offered == 8 ~ "MS",
      lowest_grade_offered == 7 & highest_grade_offered == 8 ~ "MS",
      lowest_grade_offered == 6 & highest_grade_offered == 7 ~ "MS",
      lowest_grade_offered == 6 & highest_grade_offered == 9 ~ "MS",
      lowest_grade_offered == 3 & highest_grade_offered == 8 ~ "ES/MS",
      lowest_grade_offered == 4 & highest_grade_offered == 8 ~ "MS",
      lowest_grade_offered == 7 & highest_grade_offered == 9 ~ "MS/HS",
      lowest_grade_offered == 7 & highest_grade_offered == 10 ~ "MS/HS",
      lowest_grade_offered == 6 & highest_grade_offered == 12 ~ "MS/HS",
      lowest_grade_offered == 7 & highest_grade_offered == 12 ~ "MS/HS",
      lowest_grade_offered == 8 & highest_grade_offered == 12 ~ "HS",
      is.na(lowest_grade_offered)  ~ "UG",
      lowest_grade_offered == "Ungraded" | highest_grade_offered == "Ungraded" ~ "UG",
      TRUE ~ "All Grades"))

# Rough cross-tab to compare my coding with Urban Institutes
# On the whole, I better define the overlap ranges
# in some cases they'll code "primary" if it spans into all grades; and "other" for all grades

#all_public_dir %>% 
#  tabyl(grade_band, school_level)

```

```{r NCES private grade levels, include = FALSE}

# Private Schools dataset presents low and high grade columns as AGES.  (Because, of course it does.)
# There's a table in the Word doc data lot that explains the groupings, in which I've already convered ages to grades and made rules.  
# This coding should get to the final grade bands out of the age range data.

all_pvt = all_pvt %>% 
  mutate(grade_band = 
            case_when(
              hi_grade < 12 & lo_grade > 1 ~ "ES",
              lo_grade > 12 ~ "HS",
              lo_grade < 9 & hi_grade == 12 ~ "ES/MS",
              lo_grade < 9 & hi_grade == 13 ~ "ES/MS",
              hi_grade > 13 & lo_grade == 10 ~ "MS/HS",
              hi_grade > 13 & lo_grade == 11 ~ "MS/HS",
              hi_grade > 13 & lo_grade == 12 ~ "MS/HS",
              hi_grade == 12 & lo_grade == 9 ~ "MS",
              hi_grade == 12 & lo_grade == 10 ~ "MS",
              hi_grade == 12 & lo_grade == 11 ~ "MS",
              hi_grade == 13 & lo_grade == 9 ~ "MS",
              hi_grade == 13 & lo_grade == 10 ~ "MS",
              hi_grade == 13 & lo_grade == 11 ~ "MS",
              hi_grade == 13 & lo_grade == 12 ~ "MS",
              lo_grade == 1 ~ "UG",
              TRUE ~ "All Grades"))
              
```

### Cleaning Other Variables

Each data set is appended with a new variable indicating whether the school is public or private.

**Public Schools:**

-   There is a variable of school_status -- which includes a handful of schools with the status of "Closed". **These should be cleaned out for analysis;** but keep in the data set we work with.

-   Specialized public school types (magnet, charter) are currently coded each as their own yes/no variable. These are mutually exclusive categories, so we will merge them into a single variable of "School Type" (school_type_detail) with three options: Magnet, Charter, and Public (where "Public" means non-magnet, non-charter public school).

-   The variable for Title I status is, essentially, blank to indicate "No." After cleaning symbols in the data set (footnotes in a web version, essentially), we transform NA data points into "No".

-   Data on FARM metrics are raw numbers in separate variables. We will compute a standardized ratio representing Percent of (all enrolled) Students Eligible for FARM (free or reduced meals).

*Note: Title I and FARM are the only available metrics of diversity in public school data. Racial composition is not reported by NCES at a school-by-school level (only at a district level).*

**Private Schools:**

-   Private schools report racial/ethnicity data of student body. From this, we compute a Percentage of Non-White Students as an aggregated measure.
-   Private schools need to calculate total enrollment

*Note: This is the only metric of "diversity" in the private school data. It's not remotely equivalent to Title I or FARM, but it is the metric available.*

```{r NCES other cleaning, include = FALSE}

# First we append each dataset with a new variable of its source (public, private)

# We create a new variable that distinguishes the type of public school - regular public, charter, or magnet.

# We create a simplified Title I variable

# We calculate a percentage of enrolled students eligible for FARM

all_public = all_public %>% 
  mutate(school_type = "public") %>% 
  mutate(school_type_detail = case_when(
    magnet == "Yes" ~ "magnet",
    charter == "Yes" ~ "charter",
    TRUE ~ "public")) %>% 
  mutate(title_i_school = case_when(
    title_i_status == "School is not eligible for either Title I Targeted Assistance or Schoowide program" ~ "Not Title I",
    TRUE ~ "Title I")) %>% 
  mutate(farm_pct = free_or_reduced_price_lunch/enrollment)



# Both datasets need county codes turned into county names

county_codes = read_csv("raw-data/md_county_codes.csv")

all_public = merge(all_public, county_codes, by = "county_code")

all_pvt = all_pvt %>% 
  rename(county_code = pss_county_no)

all_pvt = merge(all_pvt, county_codes, by = "county_code")


# Private Data Cleaning

# Adding "private" code
# Calculating the percentage of non-white students
# changing the case of the school names to Title Case to match the Public data

all_pvt = all_pvt %>% 
  mutate(school_type = "private") %>% 
  mutate(school_type_detail = "private") %>%
  rowwise() %>% 
  mutate(pvt_pct_nonwhite = sum(pss_indian_pct, pss_hisp_pct, pss_black_pct, pss_pacisl_pct, pss_twomore_pct)) %>% 
  mutate(pss_inst = str_to_title(pss_inst))


# The private data only has locale codes
# Those need to be changed to the same names as UI

locale_codes = read_csv("raw-data/locale_codes.csv")

all_pvt = merge(all_pvt, locale_codes, by = "pss_locale")


```

### Preparing & Merging into All Schools Data File

This table is of variables for merging; where there are gaps in one file, there isn't a match.

|                             |                      |                             |
|-----------------------------|----------------------|-----------------------------|
| all_public                  | all_pvt              | NEW                         |
| ncessch                     | pss_school_id        | school_id                   |
| school_name                 | pss_inst             | school_name                 |
| county_name                 | county_name          | county_name                 |
| street_location             | pss_address          | street                      |
| city_location               | pss_city             | city                        |
| state_location              | pss_stabb            | state                       |
| zip_location                | pss_zip5             | zip                         |
| urban_centric_locale        | urban_centric_locale | locale_label                |
| grade_band                  | grade_band           | grade_band                  |
| school_type                 | school_type          | school_type                 |
| school_type_detail          | school_type_detail   | school_type_detail          |
| teachers_fte                | pss_fte_teach        | teachers_fte                |
| enrollment                  | pss_enroll_t         | enrollment                  |
| title_i\_status             |                      | title_i\_status             |
| title_i\_school             |                      | title_i\_school             |
| free_or_reduced_price_lunch |                      | free_or_reduced_price_lunch |
| farm_pct                    |                      | farm_pct                    |
| lunch_program               |                      | lunch_program               |
|                             | pvt_pct_nonwhite     | pvt_pct_nonwhite            |
| school_status               |                      | school_status               |

Variables not on the list will be dropped in merging.

The next step is to change variable names to the "new" column, which will make merging easier.

Then we will save copies of the separate public and private files with cleaned data in the "data" folder. We will probably not use them in analysis, preferring the merged file. But it's good to have them.

```{r NCES merge and write, include = FALSE}

# Create new dataframes with renamed columns to make the merge easier.

all_public = all_public %>% 
  rename(school_id_2 = school_id,
         school_id = ncessch,
         street = street_location,
         city = city_location,
         state = state_location,
         zip = zip_location,
         locale_label = urban_centric_locale)

all_pvt = all_pvt %>% 
  rename(school_id = pss_school_id,
         school_name = pss_inst,
         street = pss_address,
         city = pss_city,
         state = pss_stabb,
         zip = pss_zip5,
         locale_label = urban_centric_locale,
         enrollment = pss_enroll_t,
         teachers_fte = pss_fte_teach) %>% 
  mutate(school_status = "Open")

write_csv(all_public, "data/all_public_cleaned.csv")

write_csv(all_pvt, "data/all_private_cleaned.csv")

## Merge the two files to get "All Schools" with just key data.

# Write this file to data for analysis in the actual report file

all_schools = merge(all_public, all_pvt, all = TRUE) %>% 
  select(school_id,
         school_name,
         county_name,
         street,
         city,
         zip,
         locale_label,
         grade_band,
         school_type,
         school_type_detail,
         teachers_fte,
         enrollment,
         title_i_status,
         title_i_school,
         free_or_reduced_price_lunch,
         farm_pct,
         lunch_program,
         pvt_pct_nonwhite,
         school_status)

write_csv(all_schools, "data/all_schools_cleaned.csv")
```

## Exploratory Descriptive Statistics

### All Maryland Schools

-   Public school data included a school_status variable, which indicated schools as "Closed" (as well as open, reopened, etc.)

-   Before any analysis, the file must be filtered to get rid of "closed" status. But we want to keep them in to aid with the ongoing set over time.

Below are frequency tables summarizing the data. These can also be used to compare to outputs from Excel-based visualizations to double-check and spot possible errors in data files, outputs, etc.

#### Public and Private

```{r tables all schools}

# The first line of code is to filter out any schools that, as of the last data run, were coded as "Closed."

all_schools = all_schools %>% 
  filter(school_status != "Closed")


# Flextable is the formatting for the tables

all_schools %>% 
  tabyl(school_type) %>% 
  flextable() %>% 
  theme_zebra()

all_schools %>% 
  tabyl(school_type_detail) %>% 
  flextable() %>% 
  theme_zebra()
```

#### By County

```{r tables all by county}
all_schools %>% 
  tabyl(county_name)%>% 
  flextable() %>% 
  theme_zebra()

all_schools %>% 
  tabyl(county_name, school_type) %>% 
  adorn_totals(where = c("row", "col"))%>% 
  flextable() %>% 
  theme_zebra()
```

#### By Locale Type

```{r tables all by locale}
all_schools %>% 
  tabyl(locale_label)%>% 
  flextable() %>% 
  theme_zebra()

all_schools %>% 
  tabyl(locale_label, school_type) %>% 
  adorn_totals(where = c("row", "col"))%>% 
  flextable() %>% 
  theme_zebra()
```

#### By Grade Band

> **Code Note**: I don't know how to order tables with text-based options to put them in the proper sequence (not alpha).

```{r tables all by grade}
all_schools %>% 
  tabyl(grade_band)%>% 
  flextable() %>% 
  theme_zebra()

all_schools %>% 
  tabyl(grade_band, school_type) %>% 
  adorn_totals(where = c("row", "col"))%>% 
  flextable() %>% 
  theme_zebra()
```

#### By SES or Racial Diversity Measures

The measures are different for public and private schools. For public schools, we have:

-   Percentage that are designated Title I
-   Several descriptives of the FARM eligibility. A histogram of the data may ultimately be better.
-   Several descriptives of the percentages of non-white students at private schools/

```{r tables all by diversity, eval = FALSE}

#### CODED NOT TO RUN BECAUSE THE VARIABLE ISN'T WORKING; SEE ABOVE

all_schools %>% 
  drop_na(title_i_school) %>% 
  tabyl(title_i_school)%>% 
  flextable() %>% 
  theme_zebra()

all_schools %>% 
  drop_na(farm_pct) %>% 
  summarize("Average % Students Eligible for FARM" = mean(farm_pct),
            "Median % Students Eligible for FARM" = median(farm_pct),
            "Percent of Schools w/ >50% FARM eligible students" = (sum(farm_pct > .50)/n()),
            "Percent of Schools w/ <25% FARM eligible students" = (sum(farm_pct < .25)/n()),
            "Percent of Schools w/ >75% FARM eligible students" = (sum(farm_pct > .75)/n()))%>% 
  flextable() %>% 
  theme_zebra()

all_schools %>% 
  drop_na(pvt_pct_nonwhite) %>% 
  summarize("Average % Enrollment of Non-White Students" = mean(pvt_pct_nonwhite),
            "Median % Enrollment of Non-White Students" = median(pvt_pct_nonwhite),
            "Percent of Schools w/ >50% enrollment is Non-White Students" = (sum(pvt_pct_nonwhite > 50)/n()),
            "Percent of Schools w/ >25% enrollment is Non-White Students" = (sum(pvt_pct_nonwhite > 25)/n()))%>% 
  flextable() %>% 
  theme_zebra()

```

### All Public Schools

MAEOE asked for data to be presented separately between public and private schools. This section contains the same descriptive tables, but separated by public/private.

#### By County

```{r tables all public by county}
# The first line of code is to filter out any schools that, as of the last data run, were coded as "Closed."

all_public = all_public %>% 
  filter(school_status != "Closed")

all_public %>% 
  tabyl(county_name)%>% 
  flextable() %>% 
  theme_zebra()

all_public %>% 
  tabyl(county_name, school_type) %>% 
  adorn_totals(where = c("row", "col"))%>% 
  flextable() %>% 
  theme_zebra()
```

#### By Locale Type

```{r tables all public by locale}
all_public %>% 
  tabyl(locale_label)%>% 
  flextable() %>% 
  theme_zebra()

all_public %>% 
  tabyl(locale_label, school_type) %>% 
  adorn_totals(where = c("row", "col"))%>% 
  flextable() %>% 
  theme_zebra()
```

#### By Grade Band

```{r tables all public by grade}
all_public %>% 
  tabyl(grade_band)%>% 
  flextable() %>% 
  theme_zebra()

all_public %>% 
  tabyl(grade_band, school_type) %>% 
  adorn_totals(where = c("row", "col"))%>% 
  flextable() %>% 
  theme_zebra()
```

### All Private Schools

MAEOE asked for data to be presented seprately between public and private schools.

#### By County

```{r tables all private by county}
all_pvt %>% 
  tabyl(county_name)%>% 
  flextable() %>% 
  theme_zebra()

all_pvt %>% 
  tabyl(county_name, school_type) %>% 
  adorn_totals(where = c("row", "col"))%>% 
  flextable() %>% 
  theme_zebra()
```

#### By Locale Type

```{r tables all private by locale}
all_pvt %>% 
  tabyl(locale_label)%>% 
  flextable() %>% 
  theme_zebra()

all_pvt %>% 
  tabyl(locale_label, school_type) %>% 
  adorn_totals(where = c("row", "col"))%>% 
  flextable() %>% 
  theme_zebra()
```

#### By Grade Band

```{r tables all private by grade}
all_pvt %>% 
  tabyl(grade_band)%>% 
  flextable() %>% 
  theme_zebra()

all_pvt %>% 
  tabyl(grade_band, school_type) %>% 
  adorn_totals(where = c("row", "col"))%>% 
  flextable() %>% 
  theme_zebra()
```
